---
title: "T05: Multiclass Logistic Regression"
subtitle: "MATH 4432 Statistical Machine Learning"
author: "WANG Zhiwei"
institute: "MATH, HKUST"
date: "2022-10-11"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = "asis",
  echo = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(xaringanthemer)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Work Sans", "300", "300i"),
  code_font_google   = google_font("IBM Plex Mono")
)

style_mono_light(
  base_color = "#003366",
  link_color = "#996600",
  text_bold_color = "#996600",
  link_decoration = "underline"
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r echo=FALSE, message=FALSE}
library(ggplot2)
```

class: inverse, center, middle

# Let's first recall what we have learned in class!

---

## Logistic regression for classification

- Training set $\mathcal{D}=\left\{\mathbf{x}_i, y_i\right\}_{i=1}^N$, where $y_i \in\{0,1\}$.

--

- Probabilistic model
$$
p(y \mid \mathbf{x}, \beta)=\operatorname{Ber}\left(y \mid \sigma\left(\beta^{\top} \mathbf{x}\right)\right)
$$
    - $\sigma(z)$ is the sigmoid/logistic/logit function.
        $$
        \sigma(z)=\frac{1}{1+\exp (-z)}=\frac{e^z}{e^z+1}
        $$
    - It maps $\mathbb{R}$ to $(0,1)$.
    
---

## Logit function
    
```{r out.width='60%', fig.align='center', echo=FALSE}
set.seed(20221011)
library(ggplot2)

x <- runif(1000, -10, 10)
y <- 1 / (1 + exp(-x))
ggplot(NULL, aes(x = x, y = y)) +
  geom_line(color = "red", size = 2) +
  xlab(NULL) +
  ylab(NULL) +
  theme_bw() +
  theme(
    text = element_text(size = 24),
    axis.title = element_text(size = 24),
    axis.text.y = element_text(size = 24),
    axis.text.x = element_text(size = 24)
    )
```

---

## Joint probability

- Recall that, the likelihood is the joint probability function of joint density function of the data.

- Here, we have independent observations $\left(\mathbf{x}_i, y_i\right), i=1, \ldots, n$, each follows the (conditional) distribution
$$
\operatorname{Pr}\left(y_i=1 \mid \mathbf{x}_i\right)=\frac{1}{1+\exp \left(-\beta^T \mathbf{x}_i\right)}=1-\operatorname{Pr}\left(y_i=0 \mid \mathbf{x}_i\right)
$$

- So, the joint probability function is
$$\prod_{i=1, \ldots, n ; y_i=1} \operatorname{Pr}\left(y_i=1 \mid \mathbf{x}_i\right) \prod_{i=1, \ldots, n ; y_i=0} \operatorname{Pr}\left(y_i=0 \mid \mathbf{x}_i\right)$$
which can be conveniently written as
$$
\prod_{i=1}^n \frac{\exp \left(y_i \beta^T \mathbf{x}_i\right)}{1+\exp \left(\beta^T \mathbf{x}_i\right)}
$$

---

## The maximum likelihood estimation

- The likelihood function is the same as the joint probability function, but viewed as a function of $\beta$. The log-likelihood function is
$$
\ell=\sum_{i=1}^n\left[y_i \beta^T x_i-\log \left(1+\exp \left(\beta^T \mathbf{x}_i\right)\right)\right]
$$

- Unlike linear regression, we can no longer write down the MLE in closed form. Instead, we need to use optimization algorithms to compute it.
    - Gradient descent
    - Newton’s method

---
class: inverse, center, middle

# Now let's go to multiclass logistic regression!

---

## A set of independent binary regressions

We now extend the two-class logistic regression approach to the setting of $K > 2$ classes.
This extension is known as multiclass logistic regression or multinomial logistic regression.

--

To do this, we first select a single class to serve as the **baseline** (why?); without loss of generality, we select the $K$-th class for this role.
Then
$$\log \frac{\operatorname{Pr}\left(Y_i=k\right)}{\operatorname{Pr}\left(Y_i=K\right)}=\boldsymbol{\beta}_k^T \mathbf{x}_i ,$$
for $k=1, \ldots, K-1$.
**Notice that the log odds between any pair of classes is linear in the features.**

--

We have introduced separate sets of regression coefficients, one for each possible outcome.
If we exponentiate both sides, and solve for the probabilities, we get
$$\operatorname{Pr}\left(Y_i=k\right)=\operatorname{Pr}\left(Y_i=K\right) e^{\beta_k^T \mathbf{x}_i} .$$
---

## Sum $K$ probabilities

Using the fact that all $K$ of the probabilities must sum to one, we find
$$\operatorname{Pr}\left(Y_i=K\right)=1-\sum_{k=1}^{K-1} \operatorname{Pr}\left(Y_i=k\right)=1-\sum_{k=1}^{K-1} \operatorname{Pr}\left(Y_i=K\right) e^{\beta_ k^T \mathbf{x}_i} \\ 
\Rightarrow \operatorname{Pr}\left(Y_i=K\right)=\frac{1}{1+\sum_{k=1}^{K-1} e^{\beta_k^T \mathbf{x}_i}} .$$
We can use this to find the other probabilities generally
$$\operatorname{Pr}\left(Y_i=k\right)=\frac{e^{\beta_k^T \mathbf{x}_i}}{1+\sum_{k=1}^{K-1} e^{\beta_k^T \mathbf{x}_i}} ,$$
where $\beta_K$ is defined to be zero.

<!-- --- -->

<!-- ## Estimating the coefficients -->


---
class: inverse, center, middle

# Good night!

Slides created via Yihui Xie's R package [**xaringan**](https://github.com/yihui/xaringan).

Theme customized via Garrick Aden-Buie's R package [**xaringanthemer**](https://github.com/gadenbuie/xaringanthemer).

Tabbed panels created via Garrick Aden-Buie's R package [**xaringanExtra**](https://github.com/gadenbuie/xaringanExtra/).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).